Statistical Model Selection for Markov Models of Biomolecular Dynamics

### 2014-04-15 paper/manuscript/manuscript.tex
manuscript round 2

Instead, we explore three alternative procedures for choosing the number of states in an MSM: cross validation, Schwarz's Bayesian information criterion (BIC)\cite{Schwartz78Estimating} and the Akaike information criterion (AIC)\cite{Akaike1974AIC}. Cross-validation attempts to directly measure the Markov model's generalization error. First, the model is parameterized by building both the state space and transition matrix on a subset of the available molecular dynamics trajectories, then the likelihood is evaluated on the left-out portion. This scheme can be repeated many times with different partitions of the data set.
\begin{equation}
  \label{eq:bic}
  \mathrm{BIC} \equiv -2\cdot \ln L + \left(\ln N \right)\cdot \kappa
\end{equation}
\begin{equation}
  \label{eq:aic}
  \mathrm{AIC} \equiv -2\cdot \ln L + \left(2\right) \cdot \kappa
\end{equation} where $L$ is the maximum likelihood, $\kappa$ is the number of free parameters, and $N$ is the number of data points.  Model selection is performed by a minimization of the criterion. The BIC is derived based on the dominant terms in the Laplace approximation to the logarithm of the Bayes factor with a vague prior, while the functional form of the AIC comes from an asymptotic approximation to the Kullback-Leibler divergence between the model and the true distribution\cite{kass1995bayes, liddle2004many}. The appearance of number of data points $N$ in \cref{eq:bic} assumes that the data is independent and identically distributed given the model, which is often poorly justified for time series. Nonetheless, the value in the AIC and BIC comes from their simple form, that they do not require leaving out portions of the available data during fitting, and acceptable performance in practice\cite{Kuha01112004}.

@@

\begin{figure}
\centering
\label{fig:ww-tica}
\includegraphics[width=6in]{figs/ww_tica.pdf}

\caption{MSMs for the Fip35WW domain built with tICA were evaluated based on four model selection criteria. The log-likelihood was evaluated on the training set (black) and a separate test set (blue), and compared with the AIC (purple) and BIC (red) penalized log-likelihoods, computed on the training set. Dashed lines are used to indicate criterion-minimizing models. The models' ten slowest relaxation timescales (shown in the right panel) converge rapidly. }
\end{figure}

@@

>>>>

Instead, we explore three alternative procedures for choosing the number of states in an MSM: cross validation, Schwarz's Bayesian information criterion (BIC)\cite{Schwartz78Estimating} and the Akaike information criterion (AIC)\cite{Akaike1974AIC}. Cross-validation attempts to directly measure the Markov model's generalization error. First, the model is parameterized by building both the state space and transition matrix on a subset of the available molecular dynamics trajectories, then the likelihood is evaluated on the left-out portion. This scheme can be repeated many times with different partitions of the data set.
\begin{equation}
  \label{eq:bic}
  \mathrm{BIC} \equiv -2\cdot \ln L + \left(\ln N \right)\cdot \kappa
\end{equation}
\begin{equation}
  \label{eq:aic}
  \mathrm{AIC} \equiv -2\cdot \ln L + \left(2\right) \cdot \kappa
\end{equation} where $L$ is the maximum likelihood, $\kappa$ is the number of free parameters, and $N$ is the number of data points.  Model selection is performed by a minimization of the criterion. The BIC is derived based on the dominant terms in the Laplace approximation to the logarithm of the Bayes factor with a vague prior, while the functional form of the AIC comes from an asymptotic approximation to the Kullback-Leibler divergence between the model and the true distribution\cite{kass1995bayes,Liddle2007Information}. The appearance of number of data points $N$ in \cref{eq:bic} assumes that the data is independent and identically distributed given the model, which is often poorly justified for time series. Nonetheless, the value in the AIC and BIC comes from their simple form, that they do not require leaving out portions of the available data during fitting, and acceptable performance in practice\cite{Kuha01112004}.

@@

\begin{figure}
\centering
\includegraphics[width=6in]{figs/ww_tica.pdf}

\caption{\label{fig:ww-tica} MSMs for the Fip35WW domain built with tICA were evaluated based on four model selection criteria. The log-likelihood was evaluated on the training set (black) and a separate test set (blue), and compared with the AIC (purple) and BIC (red) penalized log-likelihoods, computed on the training set. Dashed lines are used to indicate criterion-minimizing models. The models' ten slowest relaxation timescales (shown in the right panel) converge rapidly. }
\end{figure}

@@

\begin{figure}
\includegraphics{figs/toc.png}
\caption{TOC figure}
\end{figure}
